---
title: "01 Proteome HD data analysis"
author: "Joanna A. Krupka"
output: 
  html_document:
    theme: cosmo
    code_folding: hide
---

### Introduction  

Protome HD database uses data from  5,288 individual mass-spectrometry runs across multiple tissue-types and 294 biological conditions to infer a co-regulation map of the human proteome (Kustatscher et al., 2019). 

#### Objectives  

Analyse DDX3X interactome downloaded from Proteome HD.

### Materials and methods  

```{r eval = T, echo = F, message = F, warning = F, error = F}
library(tidyverse)
library(pheatmap)
library(biomaRt)
library(RColorBrewer)
library(msigdbr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)
library(ggrepel)

source("../utilis.R")

# Reference set
refU2932 <- read_csv("../utilis/refList/U2932_refList.csv")
refMutu <- read_csv("../utilis/refList/Mutu_refList.csv")

# Load data (downloaded from Proteome HD)
tSNE <- read_csv("tSNE.csv") %>%
  rownames_to_column("Lp") %>%
  mutate(Lp = as.integer(Lp))

DXinteractome <- read_csv("DDX3X_interactome.csv", skip = 2) %>%
  dplyr::select(-starts_with("X")) %>%
  group_by(uniprot_acc) %>%
  mutate(Lp = grep(uniprot_acc, tSNE$ProteinID)[1])

# Map to entrezID
ens <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
biomart_ensembl2name <- getBM(attributes=c('external_gene_name',
                                           'entrezgene_id'), 
                                filters = "external_gene_name", 
                                values = DXinteractome$gene_names, 
                                mart = ens)
DXinteractome <- DXinteractome %>%
  left_join(biomart_ensembl2name, by = c("gene_names"="external_gene_name"))

tSNE <- tSNE %>%
  left_join(DXinteractome) 
```

```{r eval = T, echo = F, message = F, warning = F, error = F}
# Gene ontology analysis od interacting proteins
toSave <- c()
ontologies <- c("BP","CC", "MF")

ontologyTab <- tibble()
signifList <- DXinteractome$entrezgene_id
universeList <- unique(c(refU2932$entrezgene_id, refMutu$entrezgene_id))

for (o in 1:3){
    goTab<- enrichGO(gene = as.character(signifList),
                universe = as.character(universeList),
                OrgDb = org.Hs.eg.db,
                ont = ontologies[o],
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,
                readable = T)
    
      goTab <- as.data.frame(goTab) %>%
        separate(GeneRatio, c("n", "N"), "/", remove = F) %>%
        separate(BgRatio, c("b", "B"), "/", remove = F) %>%
        mutate(n = as.numeric(n),
               N = as.numeric(N),
               b = as.numeric(b),
               B = as.numeric(B),
               EnrichmentScore = (n/N)/(b/B))  %>%
        filter(n > 10,
               b > 20)  %>% 
        arrange(desc(EnrichmentScore)) %>%
        mutate(group = paste("GO", ontologies[o]),
               sample = "Co-IP") 
          
      toSave <- goTab
      
      if(nrow(ontologyTab) == 0){
        ontologyTab <- toSave
      } else {
        ontologyTab <- ontologyTab %>%
          full_join(toSave)
      }      
  }


GO_CoIP <- ontologyTab
write_csv(ontologyTab, "ontologies.csv")

# Pathways
paths_CoIP <- enrichPathway(gene = as.character(signifList),
                universe = as.character(universeList),
                pAdjustMethod = "BH",
                pvalueCutoff = 0.1,
                qvalueCutoff = 0.2,
                readable = T)
paths_CoIP <- as.data.frame(paths_CoIP)
```

```{r eval = T, echo = F, message = F, warning = F, error = F}
# Visualise
background <- is.na(tSNE$class)

cols <- c("black", divergingPal[c(8,11,2,10)])

ggplot() +
  geom_point(aes(x = tSNE$tSNE_x_dim[background], y = tSNE$tSNE_y_dim[background]), 
             size = 0.6, alpha = 0.8, colour = "darkgrey") +
   geom_point(aes(x = tSNE$tSNE_x_dim[!background], y = tSNE$tSNE_y_dim[!background],
                  color = tSNE$class[!background]), size = 1.5) +
  theme_bw() +
  scale_colour_manual(values = cols) +
  labs(x = "tSNE dimension 1", y = "tSNE dimension 2", colour = "class") 

ggsave(file.path("plots", "tSNEfull.pdf"),
           width = 9, height = 6, useDingbats=FALSE)

ggplot() +
  geom_point(aes(x = tSNE$tSNE_x_dim[background], y = tSNE$tSNE_y_dim[background]), 
             size = 2, alpha = 0.6, colour = "darkgrey") +
   geom_point(aes(x = tSNE$tSNE_x_dim[!background], y = tSNE$tSNE_y_dim[!background],
                  color = tSNE$class[!background]), size = 4) +
   geom_text_repel(aes(x = tSNE$tSNE_x_dim[!background], y = tSNE$tSNE_y_dim[!background],
                  color = tSNE$class[!background], label = tSNE$gene_names[!background]), size = 2,
                 box.padding = 0.3) +
  theme_bw() +
  scale_colour_manual(values = cols) +
  labs(x = "tSNE dimension 1", y = "tSNE dimension 2", colour = "class") +
  lims(x = c(-65,-30), y = c(-40,60))

ggsave(file.path("plots", "tSNEzoom.pdf"),
           width = 9, height = 6, useDingbats=FALSE)
```

### Conclusions  

I found 81 proteins coregulated with DDX3X. 27 of these were components of cytosolic ribosome, 8 were components 
of core translation machinery. Most enriched GO terms of DDX3X interactome were RNP complex, mRNA metabolism, translation initiation and cytosolic ribosome.  