---
title: "01 Differential Expression"
author: "Joanna Krupka"
date: "12/11/2019"
output: html_document
---

## Introduction

Here we examine the effect of MYC in primary germinal center (GC) B-cells. We cotransduced GC B cells with MYC and BCL2. As controls, cells were transduced with either BCL2 alone or BCL6 and BCL2.

## Objectives

Perform differential expression analysis comparing MYC-BCL2 transduced GC B-cells with BCL2 alone.

## Material & Methods

```{r eval = T, echo = T, message = F, warning = F, cache = F, error = F}
library(DESeq2)
library(clusterProfiler)
library(tidyverse)
library(pheatmap)
library(biomaRt)
library(org.Hs.eg.db)
library(ReactomePA)
library(msigdbr)
library(enrichplot)

source("../utilis.R")

# Functions 
doGO <- function(signifList, universeList, name){
  
  toSave <- c()
  ontologies <- c("BP","CC", "MF")
  ontologyTab <- tibble()

  for (o in 1:3){
    goTab<- enrichGO(gene = as.character(signifList),
                universe = as.character(universeList),
                OrgDb = org.Hs.eg.db,
                ont = ontologies[o],
                pAdjustMethod = "BH",
                pvalueCutoff = 0.1,
                qvalueCutoff = 0.2,
                readable = T)
    
      goTab <- as.data.frame(goTab) %>%
        separate(GeneRatio, c("n", "N"), "/", remove = F) %>%
        separate(BgRatio, c("b", "B"), "/", remove = F) %>%
        mutate(n = as.numeric(n),
               N = as.numeric(N),
               b = as.numeric(b),
               B = as.numeric(B),
               EnrichmentScore = (n/N)/(b/B))  %>%
       # filter(n > 10,
      #         b > 20)  %>% 
        arrange(desc(EnrichmentScore)) %>%
        mutate(group = paste("GO", ontologies[o]),
               status = name,
               sample = "U2932") 
          
      toSave <- goTab
      
      if(nrow(ontologyTab) == 0){
        ontologyTab <- toSave
      } else {
        ontologyTab <- ontologyTab %>%
          full_join(toSave)
      }      
   }
  return(ontologyTab)
}

# Data
load("../utilis/expressionData/GCBcells/RNA_counts.RData")
GCBcounts <- counts$counts %>%
  as.data.frame() %>%
  rownames_to_column("ensembl") 

colnames(GCBcounts)[-1] <- splitvec(colnames(GCBcounts)[-1], "[.]", 13:15, merge = ".")

metadata <- read_csv("../utilis/expressionData/GCBcells/metadata.csv") %>%
  filter(Sample %in% colnames(GCBcounts))

# Annotations
ens <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
biomart_ensembl2name <- getBM(attributes=c('ensembl_gene_id', 
                                           'external_gene_name', 
                                           'entrezgene_id'), 
                                filters = "ensembl_gene_id", 
                                values = splitvec(GCBcounts$ensembl, "[.]", 1),
                                mart = ens)
```

## Analysis. 

### Differential expression analysis

```{r eval = T, echo = T, message = F, warning = F, cache = F, error = F}
# Low counts filtering
GCBcountsFiltered <- remove_low(GCBcounts, 128, 0.25, 1)

# BCL2 vs. MYC-BCL2
meta_selected <- metadata %>%
  filter(Condition %in% c("BCL2", "MYC"))

GCBcountsSelected <- GCBcounts %>%
  dplyr::select(ensembl, one_of(meta_selected$Sample)) %>%
  column_to_rownames("ensembl") %>%
  as.matrix()

dds <- DESeqDataSetFromMatrix(countData = GCBcountsSelected,
                                  colData = meta_selected,
                                  design = ~ Condition)
dds$Condition <- relevel(dds$Condition, ref = "BCL2")
dds <- DESeq(dds)
GCBVst <- assay(vst(dds))
resultsNames(dds)
res <- results(dds, name = "Condition_MYC_vs_BCL2")
MYCres <- as.data.frame(res) %>%
      rownames_to_column(var = "ensembl_gene_id") %>%
  mutate(ensembl_gene_id = splitvec(ensembl_gene_id, "[.]", 1)) %>%
      left_join(biomart_ensembl2name, by = "ensembl_gene_id") %>%
  na.omit()

write_csv(MYCres, "results/MYC_DESeq2.csv")

# Normalised
GCBTMM <- tmm(GCBcountsFiltered[,-1])
rownames(GCBTMM) <- GCBcountsFiltered[,1]
GCBTMM <- GCBTMM %>% 
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  mutate(ensembl_gene_id = splitvec(ensembl_gene_id, "[.]", 1))  %>%
      left_join(biomart_ensembl2name, by = "ensembl_gene_id") %>%
  na.omit() 

# VST
meta_selected <- metadata %>%
  filter(Condition %in% c("BCL2", "MYC"))

GCBcountsSelected2 <- GCBcountsFiltered %>%
  dplyr::select(ensembl, one_of(meta_selected$Sample)) %>%
  remove_rownames() %>%
  column_to_rownames("ensembl") %>%
  as.matrix()

dds <- DESeqDataSetFromMatrix(countData = GCBcountsSelected2,
                                  colData = meta_selected,
                                  design = ~ Condition)
dds$Condition <- relevel(dds$Condition, ref = "BCL2")
dds <- DESeq(dds)
GCBVst <- assay(vst(dds))

```

### GSEA  

```{r eval = T, echo = T, message = F, warning = F, cache = F, error = F}
#ExpressionSet construction
norm_counts <- GCBVst %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  gather(sample, VST, -ensembl_gene_id) %>%
  mutate(ensembl_gene_id = splitvec(ensembl_gene_id, "[.]", 1)) %>%
  left_join(biomart_ensembl2name) %>%
  group_by(sample, external_gene_name) %>%
  summarize(VST = mean(VST)) %>%
  filter(!duplicated(external_gene_name)) %>%
  na.omit() %>%
  ungroup() %>%
  spread(sample, VST) %>%
  remove_rownames() %>%
  column_to_rownames("external_gene_name")

#AnnotatedDataFrame with metadata
meta <- data.frame(labelDescription= colnames(meta_selected[,c(1,6)]),
                   row.names=colnames(meta_selected[,c(1,6)]))
pD <- meta_selected[,c(1,6)] %>% as.data.frame()
pD <- pD %>%
  arrange(Sample)
rownames(pD) <- pD$Sample
phenoData <- new("AnnotatedDataFrame", data=pD, varMetadata=meta)

#AnnotatedDataFrame with features data
features <- biomart_ensembl2name[,c(1,2)][match(rownames(norm_counts), biomart_ensembl2name$external_gene_name),]
rownames(features) <- rownames(norm_counts)
meta <- data.frame(labelDescription= colnames(features),
                       row.names=colnames(features))
featureData <- new("AnnotatedDataFrame", data=features, varMetadata=meta)

#ExpressionSet object
exampleSet <- ExpressionSet(assayData = as.matrix(norm_counts), 
                            phenoData = phenoData, 
                            featureData = featureData)

#create GSEA files
ArrayTools::createGSEAFiles(mydir = paste(getwd(), "GSEA_GCB", sep="/"), exampleSet, "Condition")
```

```{r}
rankedList <- read_csv("GSEA_TuI/ranked_gene_list_MYC_versus_BCL2.csv") %>%
  left_join(biomart_ensembl2name, by = c("NAME" = "external_gene_name")) %>%
  na.omit()

mSigH <- msigdbr(species = "Homo sapiens", category = "H")  %>% 
  dplyr::select(gs_name, entrez_gene)
mSigAll <- msigdbr(species = "Homo sapiens")  %>% 
  dplyr::select(gs_name, entrez_gene)

geneList <- rankedList$SCORE
names(geneList) <- as.character(rankedList$entrezgene_id)

MYC_H <- GSEA(geneList, TERM2GENE = mSigH, pvalueCutoff = 0.2, exponent = 1, seed = 1)
MYC_H <- setReadable(MYC_H, org.Hs.eg.db, keyType = "ENTREZID")
MYC_Htab <- MYC_H %>% as.data.frame()

geneSetID <- c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MTORC1_SIGNALING", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE")
gseaplot2(MYC_H, geneSetID = geneSetID, pvalue_table = TRUE,
          color = divergingPal[c(8,10,11)],
          rel_heights = c(1.5, 0.5, 0.5))
ggsave("plots/MYC_GSEAplot.pdf",useDingbats = F)

write_csv(MYC_Htab, "results/MYC_GSEA_Htab.csv")
write_csv(MYC_C2tab, "results/MYC_GSEA_GOtab.csv")
```

### Heatmap

```{r}
# Data
selectedSigns <- mSigAll %>% 
  filter(gs_name == "GO_RIBOSOME_BIOGENESIS") %>%
  left_join(Tu1_TMM, by = c("entrez_gene" = "entrezgene_id")) %>%
  na.omit()

metadata <- metadata %>%
  arrange(Condition)

plot_df <- selectedSigns %>%
  dplyr::select(metadata$Sample) 
rownames(plot_df) <- selectedSigns$external_gene_name

# Annptations: columns
annotation_col <- data.frame(Condition = metadata$Condition,
                             stringsAsFactors = F) 
rownames(annotation_col) <- metadata$Sample

Var1 = divergingPal[c(2,3,10)]
names(Var1) = unique(annotation_col$Condition) 

ann_colors = list(Condition = Var1) 

# Annptations: rows
annotation_row <- data.frame(RPs = case_when(
  grepl("^MRPL", selectedSigns$external_gene_name) ~ "Mitochondrial RPs",
  grepl("^MRPS", selectedSigns$external_gene_name) ~ "Mitochondrial RPs",
  grepl("^RPL", selectedSigns$external_gene_name) ~ "Cytosolic RPs",
  grepl("^RPS", selectedSigns$external_gene_name) ~ "Cytosolic RPs",
  TRUE ~ "Other"),stringsAsFactors = F) 

rownames(annotation_row) <- rownames(plot_df)

Var1 = divergingPal[c(2,3,10)]
names(Var1) = unique(annotation_col$Condition) 
Var2 = divergingPal[c(12, 10,11)]
names(Var2) = unique(annotation_row$RPs) 

ann_colors = list(Condition = Var1,
                  RPs = Var2) 

# Heatmap graph parameters
colors <- c(-4, seq(-2.99,2.99,by=0.01),4) 
palette <- rev(colorRampPalette(colors = c(divergingPal[2], "white", divergingPal[10]))(n = length(colors)))  

ph <- pheatmap(plot_df, 
         scale = "row", 
         cluster_rows = T, 
         cluster_cols = F,
         color = palette, 
         breaks = colors, 
         annotation_col = annotation_col, 
         annotation_row = annotation_row,
         annotation_colors = ann_colors,
         border_color = NA,
         show_colnames = F,
         show_rownames = F, 
         width = 5, 
         height = 5,
         cutree_cols = 2,
         filename = "plots/RibosomeBiogenesis_heatmap.pdf", 
         treeheight_row = 10,
         treeheight_col = 10)
```
