---
title: "04 MHG Subtype identification"
author: "Joanna Krupka"
date: "30 July 2019"
output: 
  html_document:
    theme: cosmo
    code_folding: show
---

## Introduction  

DLBCL is a highly heterogenous diseases which can be classified into subtypes based on gene expression profile. 
DDX3X mutation was recently noted to be enriched amongst the MHG transcriptional subtype of DLBCL (Sha et al., 2019). 

## Objectives

The aim of this analysis is to classify patients enrolled into GOYA clinical trial into DLBCL transcriptional subtypes (cell-of-origin + Molecular High Grade) and compare the frequency of DDX3X mutations between the groups. 

## Materials and methods 

```{r eval = T, echo = T, message = F, warning = F, cache = F, error = F}
library(tidyverse)
library(biomaRt)
library(pheatmap)
library(limma)
library(HGNChelper)
library(e1071)
library(matrixStats)
library(car)
library(ggpubr)
library(corrplot)

source("../utilis.R")

GOYAMETA <- "data/GOYA_metadataFull.csv"
GOYAMUT <- "../utilis/GOYA/GOYA_DDX3X_mutation.csv"
COO_SIGN <-  "../utilis/signatures/COO_DLBCL_signatureDave2018.csv"
MHG_SIGN <- "../utilis/signatures/MHG_DLBCL_signatureSha2018.csv"
BDCPACKAGE <- "../utilis/pipelines/BDC/BDC/R"
BDCDATA <- "../utilis/pipelines/BDC/BDC/data"

# Load BDC package
sapply(list.files(BDCPACKAGE, full.names = T), source, .GlobalEnv)
sapply(list.files(BDCDATA, ".rda", full.names = T), load, .GlobalEnv)
classifierGene <- Gene

# Load GOYA expression data
load("../utilis/GOYA/GOYAcountsFull.RData")
load("../utilis/GOYA/GOYA_TMM.RData")

GOYA_meta <- read_csv(GOYAMETA)

MHG_tab <- read_csv(MHG_SIGN)
COO_tab <- read_csv(COO_SIGN)

# Annotations
ens <- useMart("ensembl", dataset="hsapiens_gene_ensembl", host = "http://jul2019.archive.ensembl.org")
biomart_ensembl2name <- getBM(attributes=c('ensembl_gene_id', 
                                           'external_gene_name', 
                                           'entrezgene_id'), 
                                filters = "ensembl_gene_id", 
                                values = splitvec(countsGOYA$geneID, "[.]", 1),
                                mart = ens)

```

Classification of GOYA trial cases into four transcriptomic subtypes: ABC, GCB, Unclassified and Molecular High Grade (MHG) DLBCL was performed as previously described (Reddy et al., 2017, 2017; Sha et al., 2019). Briefly, read counts were TMM normalised, log2 transformed and z-scores were computed across the genes. The subtype score was obtained per sample by subtracting mean z-scores of ABC-signature genes (ABC score) from mean z-scores of GCB-signature genes (GCB score). Each sample was assigned to a subtype according to the following criteria: ABC - subtype score > 0.25 and GCB score < 0.75; GCB - subtype score < −0.25 and ABC score < 0.75; Unclasssified - not assigned to ABC or GCB subtype. MHG cases were identified among the GCB group using BDC R package.  

```{r eval = T, echo = T, message = F, warning = F, error = F}
# Normalize (lower low counts filtering criteria to capture all the genes in the classifier)
GOYA_filtered <- remove_low(countsGOYA, 20, 0.1) %>% na.omit()
GOYA_norm <- log2(tmm(as.matrix(GOYA_filtered[,-1] + 1)))
GOYA_norm  <- cbind(GOYA_filtered[,1], GOYA_norm)%>% as.tibble() %>%
  mutate(V1 = splitvec(V1, "[.]", 1)) %>%
  mutate_at(vars(2:554), as.numeric)
colnames(GOYA_norm)[1] <- "ensembl" 

DDX3X_expr <- GOYA_norm %>%
  dplyr::filter(ensembl == "ENSG00000215301") %>%
  gather(Run, DDX3Xexpr, -ensembl) %>%
  mutate(DDX3Xexpr = scale(DDX3Xexpr))

# 
GOYA_COO <- COO_tab  %>%
  left_join(dplyr::select(GOYA_norm, ensembl, starts_with("SRR")), 
            by = c("ensembl_ID" = "ensembl")) %>%
  gather(key, value, -name, -COO, -ensembl_ID) %>%
  group_by(ensembl_ID) %>%
  mutate(value = scale(as.numeric(value)))

MHG_tab <- MHG_tab %>% 
  left_join(biomart_ensembl2name, by = c("name" = "external_gene_name")) %>%
  dplyr::filter(name %in% GOYA_tmm$external_gene_name) 

# Scaled values for heatmap
GOYA_plotValues <- data.frame(ensembl = c(COO_tab$ensembl_ID, 
                                          MHG_tab$ensembl_gene_id),
                              name = c(COO_tab$name, MHG_tab$name)) %>%
  left_join(dplyr::select(GOYA_norm, ensembl, starts_with("SRR")), 
            by = "ensembl")

rowNames = GOYA_plotValues$name
GOYA_plotValues <- GOYA_plotValues %>%
  dplyr::select(starts_with("SRR"))

# ABC/GCB scores and RNA-Seq based classification (Reddy et al. Cell 2018)
GOYA_COOscores <- GOYA_COO %>%
  ungroup() %>%
  group_by(key, COO) %>%
  summarize(score = sum(value)/length(value)) %>%
  spread(COO, score) %>%
  mutate(RNAseqSubtypeScore = ABC-GCB) %>%
  mutate(RNAseqSubtype = case_when(
    RNAseqSubtypeScore > 0.25 ~  "ABC",
    RNAseqSubtypeScore < -0.25 ~ "GCB",
    TRUE ~ "Unclassified"
  )) %>%
  left_join(DDX3X_expr, by = c("key" ="Run"))

GCBsamples <- GOYA_COOscores$key[GOYA_COOscores$RNAseqSubtype == "GCB"]

# MHG classification (Care M. paper)
testData <- GOYA_norm[,GCBsamples]
testDataGene <- biomart_ensembl2name$external_gene_name[match(GOYA_norm$ensembl, biomart_ensembl2name$ensembl_gene_id)]

pred = Classify(testData, testDataGene,  classifierGene="Gene26")
mhgPred <- data.frame(Run = colnames(testData),
                      pred = pred$prediction)

SubtypesOrder <- c("MHG", "GCB", "Unclassified", "ABC")
GOYA_COOscores <- GOYA_COOscores %>%
  left_join(mhgPred, by =c("key" = "Run")) %>%
  mutate(PredSubtype = case_when(
    pred.class == "BL" ~ "MHG",
    RNAseqSubtypeScore > 0.25 ~  "ABC",
    RNAseqSubtypeScore < -0.25 ~ "GCB",
    TRUE ~ "Unclassified"
  )) %>%
  left_join(dplyr::select(GOYA_meta, key, sex, DDX3X_status)) %>%
  arrange(match(PredSubtype, SubtypesOrder))

#write_csv(GOYA_COOscores, "../utilis/GOYA/GOYA_metadataUpdatedSuptypes.csv")

# Barblot
plot_df <- table(GOYA_COOscores$PredSubtype) %>%
  as.data.frame() %>%
  mutate(Subtype = factor(Var1, levels = c("MHG", "GCB", "Unclassified", "ABC")))

ggplot(plot_df, aes(x = Subtype, y = Freq, fill = Subtype)) +
  geom_bar(stat = "identity") +
  nature_barplot()+
  coord_flip() +
  scale_fill_manual(values = rev(divergingPal[c(8,9,10,11)]))

ggsave("plots/GOYA_Subtype.pdf", width = 5, height = 2)

# Heatmap  
plot_df <- GOYA_plotValues[,GOYA_COOscores$key] %>%
  distinct()

## Annotations
annotation_col <- data.frame(Subtype = GOYA_COOscores$PredSubtype,
                             DDX3Xmut = GOYA_COOscores$DDX3X_status,
                             DDX3Xexpr = round(GOYA_COOscores$DDX3Xexpr, digits = 1),
                             stringsAsFactors = F) %>%
  mutate(DDX3Xexpr = case_when(
    DDX3Xexpr > 2.5  ~  2.5,
    DDX3Xexpr < -2.5 ~ -2.5,
    TRUE ~ DDX3Xexpr),
    DDX3Xmut = case_when(
    DDX3Xmut == "MUT_WhiteList"      ~ "Mutated",
    DDX3Xmut == "MUT_Valid_Helicase" ~ "Mutated",
    DDX3Xmut == "LowExpr"         ~ "Not expressed",
    TRUE                          ~ "Wild-type"))

rownames(annotation_col) <- GOYA_COOscores$key

Var1 = brewer.pal(11, "Spectral")[c(11,10,9,8)]
names(Var1) = unique(annotation_col$Subtype) 
Var2 = c(brewer.pal(11, "Spectral")[c(2,10)], "grey")[c(1,3,2)]
names(Var2) = unique(annotation_col$DDX3Xmut) 
Var4 = rev(colorRampPalette(brewer.pal(11, "RdBu")[c(1,2,3,8,9,10)])(length(unique(annotation_col$DDX3Xexpr))))
names(Var4) = unique(annotation_col$DDX3Xexpr) 

ann_colors = list(Subtype = Var1,
                  DDX3Xmut = Var2,
                  DDX3Xexpr = Var4) 

## Heatmap graph parameters
colors <- c(-3, seq(-2.5,2.5,by=0.01),3) 
col <- brewer.pal(5,"RdBu")[c(1,5)]
palette <- rev(colorRampPalette(c(col[1], "white", col[2]))(n = length(colors)))
#palette <- rev(colorRampPalette(brewer.pal(11,"RdBu"))(n = length(colors)))

dih <- pheatmap(as.matrix(plot_df),
                scale = "row",  #scale by row
                color = palette, #colors of the heatmap
                breaks = colors, #spread of colors
                cluster_cols = F, 
                cluster_rows = F,
                labels_row = rowNames,
                show_colnames = F,
                show_rownames = T,
                fontsize_row = 6, 
                cellwidth = 0.4, 
                cellheight = 5,
                clustering_method = "complete", #clustering parameter (see hclust function)
                annotation_col = annotation_col, 
                annotation_colors = ann_colors,
                main = "GOYA",
                #filename = paste0("plots/Heatmap_GOYA_DXmut_SubtypeUpdtaed.pdf"),
                border_color = NA,
                legend = T,
                annotation_legend = T)
```

### Statistical analysis of DDX3X mutations frequency in DLBCL subtypes (GOYA)  

```{r eval = T, echo = T, message = F, warning = F, error = F}
# Clean the data  
GOYA_COOscores <- GOYA_COOscores %>%
  mutate(Subtype = factor(PredSubtype, levels = SubtypesOrder),
         DDX3Xmut = case_when(
           DDX3X_status == "MUT_WhiteList"      ~ "Mutated",
           DDX3X_status == "MUT_Valid_Helicase" ~ "Mutated",
           DDX3X_status == "LowExpr"            ~ "Not expressed",
           TRUE                             ~ "Wild-type"))

# Visualise DDX3X mutation distribution
freqTab <- table(GOYA_COOscores[,13:14]) %>%
  as.data.frame() 

# Correct on one mutation that should be a wt
freqTab[2,3] <- 16
freqTab[10,3] <- 181

freqTab <- freqTab %>%
  group_by(Subtype) %>%
  mutate(FreqNorm = Freq/sum(Freq),
         DDX3Xmut = factor(DDX3Xmut, levels = c( "Wild-type", "Not expressed", "Mutated")))

ggplot(freqTab) +
  geom_bar(aes(x = Subtype, y = FreqNorm, fill = DDX3Xmut), stat = "identity") +
  nature_barplot() +
  coord_flip() +
  scale_fill_manual(values = c(brewer.pal(11, "Spectral")[c(2,10)], "grey")[c(2,3,1)]) +
  labs(y = "Frequency", x = "", fill = "DDX3X status")

ggsave("plots/DXmut_Subtype_barplot.pdf", width = 7, height = 2)

# Chi-square test
freqTab <- freqTab %>%
  dplyr::select(Subtype, DDX3Xmut, Freq) %>%
  spread(DDX3Xmut, Freq) %>%
  column_to_rownames("Subtype") 

chisq <- chisq.test(freqTab)
chisq
chisq$observed
chisq$expected

# Pearson's residuals
col <- brewer.pal(5,"RdBu")[c(1,5)]
palette <- rev(colorRampPalette(c(col[1], "white", col[2]))(100))

pdf("plots/DXmut_Subtype_PearsonRes.pdf")
corrplot(chisq$residuals, 
         is.cor = FALSE, 
         method = "color",
         col = palette,
         tl.col = "black", 
         cl.lim = c(-2,4),
         cl.cex = 0.8)
dev.off()

# Contribution
contrib <- 100*chisq$residuals^2/chisq$statistic
corrplot(contrib, is.cor = FALSE)
```

### Statistical analysis of DDX3X mutations frequency in DLBCL subtypes (REMODEL-B)  

```{r eval = T, echo = T, message = F, warning = F, error = F}
# Load data
remodelBdf <- read_csv("data/REMODELB_mutFreq.csv") %>%
  filter(Study != "GOYA" & Subtype != "Unclassified") %>%
  dplyr::select(Subtype, Mutated, Wild_type) %>%
  gather(DDX3X_status, Freq, -Subtype) %>%
  group_by(Subtype) %>%
  mutate(FreqNorm = Freq/sum(Freq),
         DDX3Xmut = factor(DDX3X_status, levels = c( "Wild_type", "Mutated"))) %>%
  ungroup() %>%
  mutate(Subtype = factor(Subtype, levels = c("MHG", "GCB", "ABC")))

#Barplot  
ggplot(remodelBdf) +
  geom_bar(aes(x = Subtype, y = FreqNorm, fill = DDX3Xmut), stat = "identity") +
  nature_barplot() +
  coord_flip() +
  scale_fill_manual(values = c(brewer.pal(11, "Spectral")[c(10,2)])) +
  labs(y = "Frequency", x = "", fill = "DDX3X status")

ggsave("plots/DXmut_SubtypeREMODEL_barplot.pdf", width = 5.5, height = 1.5)

remodelBdf <- read_csv("data/REMODELB_mutFreq.csv") %>%
  filter(Study != "GOYA" & Subtype != "Unclassified") %>%
  dplyr::select(Subtype, Mutated, Wild_type) %>%
  column_to_rownames("Subtype")
  
chisq <- chisq.test(remodelBdf)
chisq
chisq$observed
chisq$expected

# Pearson's residuals
col <- brewer.pal(5,"RdBu")[c(1,5)]
palette <- rev(colorRampPalette(c(col[1], "white", col[2]))(100))

pdf("plots/DXmut_Subtype_PearsonRes.pdf")
corrplot(chisq$residuals, 
         is.cor = FALSE, 
         method = "color",
         col = palette,
         tl.col = "black", 
         cl.lim = c(-2,4),
         cl.cex = 0.8)
dev.off()
```

## Conclusion

DDX3X mutation is enriched in in MHG DLBCL subtype. 
