---
title: "05 Meta analysis"
author: "Joanna A. Krupka"
date: "May 6, 2020"
output: 
  html_document:
    theme: cosmo
    code_folding: show
---

## Introduction  

DDX3X is known to escape chromosome X inactivation and is commonly mutated in male-skewed cancers. 

## Objectives  

The aim of this meta analysis is to compare the frequency of DDX3X mutation between males and females  

## Materials and methods  

```{r eval = T, echo = T, message = F, warning = F, error = F}
library(tidyverse)
library(knitr)
library(meta)

source("../utilis.R")

metaTab <- read_csv("data/DXsexDistrib.csv") %>%
  na.omit()
```

### Analysis  

#### All studies

```{r eval = T, echo = T, message = F, warning = F, error = F}
m.bin <- metabin(Ee,
                 Ne,
                 Ec,
                 Nc,
                 data = metaTab,
                 studlab = paste(Author),
                 comb.fixed = FALSE,
                 comb.random = TRUE,
                 method.tau = "SJ",
                 hakn = TRUE,
                 prediction = TRUE,
                 incr = 0.1,
                 sm = "RR")

m.bin

# Visualisation
labbe.metabin(x = m.bin,
              bg = "#3288BD",
              studlab = TRUE,
              col.random = "#D53E4F")
forest(m.bin, comb.random=FALSE, hetstat=FALSE,text.fixed="MH estimate")




# Save plots
pdf("plots/metaAnalysisDotPlot_updated.pdf", useDingbats = F) 
labbe.metabin(x = m.bin,
              bg = "#3288BD",
              studlab = TRUE,
              col.random = "#D53E4F")
dev.off() 

pdf("plots/metaAnalysisForrestPlot_updated.pdf", useDingbats = F) 
forest(m.bin, comb.random=FALSE, hetstat=FALSE,text.fixed="MH estimate")
dev.off() 
```

### Conclusions  

There is statistically significant relationship between having DDX3X mutation and being male. 

