---
title: "01 Analysis of Co-IP + SILAC of endogenous DDX3X"
author: "Joanna A. Krupka"
date: "May 6, 2020"
output: 
  html_document:
    theme: cosmo
    code_folding: show
---

### Background  

The aim of this analysis is to establish the protein interactome of endogenous DDX3X 
in human lymphoid cell lines: U2932 (female, DDX3X WT) and Mutu (male, DDX3X mutant). 

### Objectives   

1. Clean the data: unify redundant protein group, select enriched proteins for downstream analysis  
2. Perform Gene Ontology and pathway analysis (Reactome)  
3. Visualise the results for both cell lines: Venn Diagram + scatter plot 

### Materials and methods  

```{r eval = T, echo = F, message = F, warning = F, error = F}
library(tidyverse)
library(magrittr)
library(clusterProfiler)
library(msigdbr)
library(org.Hs.eg.db)
library(biomaRt)
library(limma)
library(VennDiagram)
library(ggrepel)
library(ReactomePA)

source("../utilis.R")

# FUNCTIONS
# Function to parse multiple gene names. 
unify <- function(x){
 paste(unique(toupper(unlist(strsplit(x, ";")))), collapse = ";")
}

# Wrapper function for GO analysis
doGO <- function(genes, background, categories){
  
  resultsTab <- data.frame()
  for (c in 1:length(categories)){
      msigDf <- msigdbr(species = "Homo sapiens", category = categories[c]) %>%
        dplyr::select(gs_name, entrez_gene) 
      mSigRes <- enricher(as.character(genes), TERM2GENE=msigDf, 
                          universe = background, pvalueCutoff = 0.1)
      mSigResRead <- DOSE::setReadable(mSigRes, org.Hs.eg.db, keyType = "ENTREZID")
      
      mSigResTab <- as.data.frame(mSigResRead) %>%
        separate(GeneRatio, c("n", "N"), "/", remove = F) %>%
        separate(BgRatio, c("b", "B"), "/", remove = F) %>%
        mutate(EnrichmentN = as.numeric(n)/as.numeric(b),
               Category = categories[c]) %>%
        arrange(desc(EnrichmentN)) %>%
        mutate(IDs = gsub("_", " ", ID)) %>%
        mutate(IDs = tools::toTitleCase(tolower(as.character(IDs))),
               IDs = gsub("Go", "GO", IDs),
               IDs = gsub("Rna", "RNA", IDs),
               IDs = gsub("Mrna", "mRNA", IDs),
               IDs = gsub("Dna", "DNA", IDs),
               IDs = gsub("Reactome", "REACTOME", IDs),
               IDs = gsub("Utr", "UTR", IDs),
               IDs = gsub("Kegg", "KEGG", IDs),
               IDs = gsub("Poly a", "Poly-A", IDs),
               IDs = factor(IDs, levels = IDs[order(p.adjust)]),
               N = as.numeric(N),
               n = as.numeric(n),
               b = as.numeric(b),
               B = as.numeric(B))
      
        if(nrow(resultsTab) == 0){
          resultsTab <- mSigResTab
        } else {
          resultsTab <- rbind(resultsTab, mSigResTab)
        }
      
  }
  return(resultsTab)
}


# SILAC data (Max Quant output)
MUTU_DATA <-  "data/12092016_Mutu_DDX3X IP.csv"
U2932_DATA <- "data/T_Oellerich_08112016_U2932.csv"
```

### Analysis  

#### Load and organise the results selecting top enriched proteins for further analysis  

```{r eval = T, echo = F, message = F, warning = F, error = F}
# Load Mutu
mutu.tab <- read_csv(MUTU_DATA) %>%
  dplyr::select(`Gene names`, `Ratio H/L`, `Ratio H/L normalized`, Intensity) %>%
  dplyr::rename(Mutu_HL_Ratio = `Ratio H/L normalized`,
                Mutu_Intensity = Intensity) %>%
  group_by(`Gene names`) %>%
  mutate(Gene_names = unify(`Gene names`),
         MutuSignif = Mutu_HL_Ratio > 7) %>%
  ungroup() %>%
  dplyr::select(Gene_names, Mutu_HL_Ratio, Mutu_Intensity, MutuSignif) %>%
  mutate(Gene_names_rep = splitvec(Gene_names, ";", 1))

# Enrichment threshold
ggdensity(mutu.tab, x = "Mutu_HL_Ratio", rug = TRUE,
   color = "MutuSignif", fill = "MutuSignif",
   palette = divergingPal[c(12,10)]) +
  geom_vline(xintercept = 7, linetype = "longdash")
ggsave(file.path("plots", paste("QC_Mutu_HL_Ratio.pdf", sep = "_")),
           width = 5, height = 4)

# Load U2932 
U2392.tab <- read_csv(U2932_DATA) %>%
  dplyr::select(`Gene names`, `Ratio H/L`, `Ratio H/L normalized`, Intensity) %>%
  dplyr::rename(U2932_HL_Ratio = `Ratio H/L normalized`,
                U2932_Intensity = Intensity) %>%
  group_by(`Gene names`) %>%
  mutate(Gene_names = unify(`Gene names`),
         U2932Signif = U2932_HL_Ratio > 7) %>%
  ungroup() %>%
  dplyr::select(Gene_names, U2932_HL_Ratio, U2932_Intensity, U2932Signif) %>%
  mutate(Gene_names_rep = splitvec(Gene_names, ";", 1))

# Enrichment threshold
ggdensity(U2392.tab, x = "U2932_HL_Ratio", rug = TRUE,
   color = "U2932Signif", fill = "U2932Signif",
   palette = divergingPal[c(12,11)]) +
  geom_vline(xintercept = 4, linetype = "longdash")
ggsave(file.path("plots", paste("QC_U2932_HL_Ratio.pdf", sep = "_")),
           width = 5, height = 4)

# Combine Mutu and U2932 results into one table, add entrezgene IDs
# Extract significant genes
combined.raw <- mutu.tab %>%
  full_join(U2392.tab, by = "Gene_names_rep")

# Add annotations 
ens <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
biomart_ensembl2name <- getBM(attributes=c('external_gene_name',
                                           'entrezgene_id'), 
                                filters = "external_gene_name", 
                                values = combined.raw$Gene_names_rep, 
                                mart = ens)

# Use HUGO aproved gene symbols
hgnc <- read.delim("../utilis/signatures/HUGO_annotations.txt")

combined.raw <- combined.raw %>% 
  left_join(biomart_ensembl2name, by = c("Gene_names_rep" = "external_gene_name")) %>%
  group_by(Gene_names_rep) %>%
  mutate(entrezgene_id = entrezgene_id,
         ApprovedSymbol = case_when(
           is.na(entrezgene_id) ~ as.character(hgnc$Approved.symbol[grepl(Gene_names_rep,
                                                                          hgnc$Previous.symbol)][1]),
           !is.na(entrezgene_id)  ~ Gene_names_rep)) %>%
  ungroup() %>% 
  mutate(Gene_names_rep = case_when(
    !is.na(ApprovedSymbol) ~ ApprovedSymbol,
    is.na(ApprovedSymbol)  ~ Gene_names_rep)) %>%
  dplyr::select(-entrezgene_id) %>%
  distinct()

# Correct annotations based on HUGO
biomart_ensembl2name <- getBM(attributes=c('external_gene_name',
                                           'entrezgene_id'), 
                                filters = "external_gene_name", 
                                values = combined.raw$Gene_names_rep, 
                                mart = ens)

combined.raw <- combined.raw %>%
    left_join(biomart_ensembl2name, by = c("Gene_names_rep" = "external_gene_name"))

# Save for manual curation (removing redundant entries manually)
write_csv(combined.raw, "results/combined_results_raw.csv")
```

Manual curation involved collapsing redundant protein groups into consensus groups 
that were decided based on the H/L ratio in both cell lines (group with higher H/L 
ratio considered as dominant). 

```{r eval = T, echo = F, message = F, warning = F, error = F}
# Load curated, rank scores
combined.clean <- read_csv("results/combined_results_clean.csv")

# Select proteins enriched in at least one cell line of further analysis
signifTabSelected <- combined.clean %>%
  filter(U2932Signif | MutuSignif) %>%
  distinct() 
```

#### Gene ontology analysis

In order to identify enriched gene sets in DDX3X interacting proteins, I will perform 
a gene ontology analysis. As background I will use a list of genes expressed at mRNA level
in each cell line (lists obtained from RNA-Seq experiment after filtering out genes with
low expression as described in methods specyfying RNA-Seq analysis). 

```{r eval = T, echo = F, message = F, warning = F, error = F}
toSave <- c()
ontologies <- c("BP","CC", "MF")
ontologyTab <- tibble()

# Background set (load genes expressed in analysed cell lines)
refU2932 <- read_csv("../utilis/refList/U2932_refList.csv")
refMutu <- read_csv("../utilis/refList/Mutu_refList.csv")

# Create lists
universeList <- as.character(unique(c(refU2932$entrezgene_id, refMutu$entrezgene_id)))
signifList <- signifTabSelected$entrezgene_id

# GO
for (o in 1:3){
    toSave<- enrichGO(gene = as.character(signifList),
                universe = as.character(universeList),
                OrgDb = org.Hs.eg.db,
                ont = ontologies[o],
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,
                readable = T) %>% 
      as.data.frame() %>%
      separate(GeneRatio, c("n", "N"), "/", remove = F) %>%
      separate(BgRatio, c("b", "B"), "/", remove = F) %>%
      mutate(n = as.numeric(n),
             N = as.numeric(N),
             b = as.numeric(b),
             B = as.numeric(B),
             EnrichmentScore = (n/N)/(b/B))  %>%
      arrange(desc(EnrichmentScore)) %>%
      mutate(group = paste("GO", ontologies[o]),
               sample = "Co-IP") 
      
      if(nrow(ontologyTab) == 0){
        ontologyTab <- toSave
      } else {
        ontologyTab <- ontologyTab %>%
          full_join(toSave)
      }      
}

# Reactome
pathways <- enrichPathway(gene = as.character(signifList),
              universe = as.character(universeList), 
              organism = "human", readable = T)
pathwaysTab <- as.data.frame(pathways)

write_csv(ontologyTab, "results/GOanalysis_full.csv")
write_csv(pathwaysTab, "results/Reactome_full.csv")

# Barplot
plot_df <- ontologyTab %>%
  group_by(group) %>%
  mutate(rank = rank(-EnrichmentScore)) %>%
  filter(rank < 5) %>%
  ungroup() %>%
  arrange(group, p.adjust) %>%
  mutate(Description = factor(Description, levels = Description[order(p.adjust, decreasing = T)]),
         rank = factor(as.character(1:length(Description)), 
                       levels = as.character(1:length(Description))))

# Barplot
ggplot(plot_df, aes(x = Description, y = -log10(p.adjust))) +
  geom_bar(aes(fill = rank), stat = "identity") +
  geom_text(aes(label = paste(n, b, sep = "/")), 
            position = position_identity(), hjust = 1.2, 
            colour = "white") +
  facet_grid(group~., space = "free", scales = "free") +
  coord_flip() +
  nature_barplot() +
  labs(y =  bquote(~-Log[10]~'(FDR)'), x = "", colour = "") +
  theme(legend.position = "None") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(11, "Spectral")[11:9])(14)) 

ggsave(file.path("plots", paste("GO_atLeast1in3.pdf", sep = "_")),
           width = 9, height = 4)
```

#### Examining the overlap between Mutu and U2932

Venn Diagram

```{r eval = T, echo = F, message = F, warning = F, error = F}
# Venn Diagram
U2932 <- na.omit(signifTabSelected$Id[signifTabSelected$U2932Signif])
Mutu <- na.omit(signifTabSelected$Id[signifTabSelected$MutuSignif])

## Venn 1
venn.diagram(
        x = list(U2932, Mutu),
        category.names = c("U2932" , "Mutu"),
        filename = 'plots/VennDiagram.png',
        output=TRUE,
        
        # Output features
        imagetype="png" ,
        height = 500 , 
        width = 500 , 
        resolution = 300,
        compression = "lzw",
        
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = divergingPal[c(4,9)],
        
        # Numbers
        cex = .6,
        fontface = "bold",
        fontfamily = "sans",
        
        cat.cex = 0.6,
        cat.fontface = "bold",
        cat.fontfamily = "sans",
        cat.default.pos = "outer")
```

Dot plot 

```{r eval = T, echo = F, message = F, warning = F, error = F}
# Annotate 
plot_df <- combined.clean %>%
  mutate(status = case_when(
    MutuSignif & U2932Signif ~ "Both",
    MutuSignif               ~ "Mutu",
    U2932Signif              ~ "U2932",
    TRUE ~ "Not Enriched"
  ))

select <- plot_df$status == "Both"

ggplot() +
  geom_point(aes(x = plot_df$Mutu_HL_Ratio, y =  plot_df$U2932_HL_Ratio, 
                 color = plot_df$status)) +
   geom_text_repel(aes(x =  plot_df$Mutu_HL_Ratio[select], 
                       y =  plot_df$U2932_HL_Ratio[select], 
                       label = plot_df$Id[select]),
                   size = 3,
                   box.padding = 0.5) +
  nature_point() +
  lims(x = c(0,50), y = c(0,50)) +
  geom_hline(yintercept = 4, color = divergingPal[2], linetype = "longdash", alpha = 0.5) +
  geom_vline(xintercept = 7, color = divergingPal[2], linetype = "longdash", alpha = 0.5) +
  scale_x_log10(breaks = c(0.1,1,10)) +
  scale_y_log10(breaks = c(0.1,1,10)) +
  labs(x = "Co-IP Mutu (Ratio H/L)", y = "Co-IP U2932 (Ratio H/L) ",
       color = "Status") +
  scale_color_manual(values = c(divergingPal[c(2,10)], "grey", divergingPal[c(11)]))
  
ggsave(file.path("plots", paste("CoIP_MutuVsU2932.pdf", sep = "_")),
           width = 8, height = 5, useDingbats=FALSE)
```
